Supplementary material

### 2022-02-02 submission/manuscript.rmd
co-author comments

In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
logistic regression model with 6 baseline covariates, i.e. age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk. A constant effect of treatment
was included in the model. When deriving risk predictions for individuals we set
the treatment indicator to 0. More information on model development can be found
in the supplement.

@@

We used the risk linear predictor to fit the proposed methods under study for
individualizing absolute benefit predictions. All methods predicted increasing
benefits for patients with higher baseline risk predictions, but the fitted
patterns were somewhat different. The adaptive approach selected the model with
RCS smoothing with 4 knots. However, for very low baseline risk the decreasing
predicted benefit with increasing risk may be somewhat too flexible. The more
robust models, such as the linear interaction model and the model with RCS
smoothing (3 knots), gave very similar benefit predictions. These models
followed the evolution of the stratified estimates very closely and may
therefore be preferable for use in clinical practice. The linear interaction
model had somewhat lower AIC compared to the model with RCS smoothing (3 knots),
slightly better cross-validated discrimination (c-for-benefit
`r round(discriminationPerformance[2, ], 3)` vs
`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
`r round(calibrationPerformance[3, ], 4)`).

@@

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. Also, we considered a small number of
scenarios with true treatment-covariate interactions, in which our main
conclusions remained the same (Supplement, XX). Future simulation studies could
explore the effect of more extensive deviations from risk-based treatment
effects.

@@

In conclusion, the linear interaction approach is a viable option with smaller
sample sizes and/or moderately performing risk prediction models if we consider
a non-constant relative treatment effect plausible. RCS-3 is a better option
when non-monotonic deviations from a constant relative treatment effect and/or
substantial treatment-related harms are anticipated. With larger sample size, an
adaptive approach -- selecting the method with optimal AIC -- can also be
considered as an automated alternative.

>>>>

In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
logistic regression model with 6 baseline covariates, i.e. age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk. A constant effect of treatment
was included in the model. When deriving risk predictions for individuals we set
the treatment indicator to 0. More information on model development can be found
in the supplement (Section XX).

@@

We used the risk linear predictor to fit a constant treatment effect model, a
linear interaction model and a RCS-3 model individualizing absolute benefit
predictions. RCS-4 and RCS-5 models were excluded. In our simulations these
methods were always outperformed by the simpler approaches and were often
overfitted. Finally, an adaptive approach with only the 3 candidate models was
also applied.

All considered methods provided similar fits, predicting increasing benefits for
patients with higher baseline risk predictions. All models followed the
evolution of the stratified estimates very closely. The adaptive approach based
on AIC selected the constant treatment effect model. The constant treatment
effect model had somewhat lower AIC compared to the linear interaction model
slightly worse cross-validated discrimination (c-for-benefit
`r round(discriminationPerformance[1, ], 3)` vs
`r round(discriminationPerformance[2, ], 3)`) and better cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[1, ], 4)` vs
`r round(calibrationPerformance[2, ], 4)`). In conclusion, a simpler constant
treatment effect model is adequate for predicting absolute 30-day mortality
benefits of treatment with tPA in patients with with acute MI.

@@

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our conclusions
remained unchanged. Even though the average error rates increased for all the
considered methods, due to the miss-specification of the outcome model, the
linear interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased results,
espcially in the presence of moderate or strong treatment-related harms. All the
results of these simulations can be found in Supplement, Section XX. Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

@@

In conclusion, the linear interaction approach is a viable option with smaller
sample sizes and/or moderately performing risk prediction models if we consider
a non-constant relative treatment effect plausible. RCS-3 is a better option
when non-monotonic deviations from a constant relative treatment effect and/or
substantial treatment-related harms are anticipated. Increasing the complexity
of the RCS models by increasing the number of knots does not translate to
improved benefit prediction. Using AIC for model selection among the constant
treatment effect, the linear interaction and RCS-3 model is a viable option,
especially with larger sample size.

### 2022-02-02 submission/manuscript.tex
co-author comments

In line with previous analyses {[}13,14{]}, we fitted a logistic
regression model with 6 baseline covariates, i.e.~age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and
the location of MI, to predict 30-day mortality risk. A constant effect
of treatment was included in the model. When deriving risk predictions
for individuals we set the treatment indicator to 0. More information on
model development can be found in the supplement.

We used the risk linear predictor to fit the proposed methods under
study for individualizing absolute benefit predictions. All methods
predicted increasing benefits for patients with higher baseline risk
predictions, but the fitted patterns were somewhat different. The
adaptive approach selected the model with RCS smoothing with 4 knots.
However, for very low baseline risk the decreasing predicted benefit
with increasing risk may be somewhat too flexible. The more robust
models, such as the linear interaction model and the model with RCS
smoothing (3 knots), gave very similar benefit predictions. These models
followed the evolution of the stratified estimates very closely and may
therefore be preferable for use in clinical practice. The linear
interaction model had somewhat lower AIC compared to the model with RCS
smoothing (3 knots), slightly better cross-validated discrimination
(c-for-benefit 0.526 vs 0.525) and quite similar cross-validated
calibration (ICI-for benefit 0.0115 vs 0.0117).

@@

Risk-based approaches to predictive HTE estimate treatment benefit as a
function of baseline risk. A limitation of our study is that we assumed
treatment benefit to be a function of baseline risk in the majority of
the simulation scenarios. We also considered constant moderate and
strong treatment-related harms, applied on the absolute scale to expand
the range of scenarios in line with previous work {[}{\textbf{???}}{]}.
Also, we considered a small number of scenarios with true
treatment-covariate interactions, in which our main conclusions remained
the same (Supplement, XX). Future simulation studies could explore the
effect of more extensive deviations from risk-based treatment effects.

In our simulations we only focused on risk-based methods, using baseline
risk as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}15--18{]}. Many of these methods rely on
incorporating treatment-covariate interactions in the prediction of
benefit. An important caveat of such approaches is that they may be
prone to overfitting, thus exaggerating the magnitude of the predicted
benefits. In a wide range of simulation settings, a simpler risk
modeling approach was consistently better calibrated for benefit
compared to more complex treatment effect modelling approaches {[}5{]}.
Similarly, when SYNTAX score II, a model developed for identifying
patients with complex coronary artery disease that benefit more from
percutaneous coronary intervention or from coronary artery bypass
grafting was redeveloped using fewer treatment-covariate interactions
had better external performance compared to its predecessor{[}19,20{]}.

In conclusion, the linear interaction approach is a viable option with
smaller sample sizes and/or moderately performing risk prediction models
if we consider a non-constant relative treatment effect plausible. RCS-3
is a better option when non-monotonic deviations from a constant
relative treatment effect and/or substantial treatment-related harms are
anticipated. With larger sample size, an adaptive approach -- selecting
the method with optimal AIC -- can also be considered as an automated
alternative.

@@

\leavevmode\hypertarget{ref-Athey2019}{}%
{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.

\leavevmode\hypertarget{ref-Lu2018}{}%
{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
treatment effect in observational data using random forest methods.
Journal of Computational and Graphical Statistics 2018;27:209--19.
\url{https://doi.org/10.1080/10618600.2017.1356325}.

\leavevmode\hypertarget{ref-Wager2018}{}%
{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
treatment effects using random forests. Journal of the American
Statistical Association 2018;113:1228--42.
\url{https://doi.org/10.1080/01621459.2017.1319839}.

\leavevmode\hypertarget{ref-powers2018some}{}%
{[}18{]} Powers S, Qian J, Jung K, Schuler A, Shah NH, Hastie T, et al.
Some methods for heterogeneous treatment effect estimation in high
dimensions. Statistics in Medicine 2018;37:1767--87.

\leavevmode\hypertarget{ref-farooq2013anatomical}{}%
{[}19{]} Farooq V, Van Klaveren D, Steyerberg EW, Meliga E, Vergouwe Y,
Chieffo A, et al. Anatomical and clinical characteristics to guide
decision making between coronary artery bypass surgery and percutaneous
coronary intervention for individual patients: Development and
validation of syntax score ii. The Lancet 2013;381:639--50.

\leavevmode\hypertarget{ref-takahashi2020redevelopment}{}%
{[}20{]} Takahashi K, Serruys PW, Fuster V, Farkouh ME, Spertus JA,
Cohen DJ, et al. Redevelopment and validation of the syntax score ii to
individualise decision making between percutaneous and surgical
revascularisation in patients with complex coronary artery disease:
Secondary analysis of the multicentre randomised controlled syntaxes
trial with external cohort validation. The Lancet 2020;396:1399--412.
\end{cslreferences}

>>>>

In line with previous analyses {[}13,14{]}, we fitted a logistic
regression model with 6 baseline covariates, i.e.~age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and
the location of MI, to predict 30-day mortality risk. A constant effect
of treatment was included in the model. When deriving risk predictions
for individuals we set the treatment indicator to 0. More information on
model development can be found in the supplement (Section XX).

We used the risk linear predictor to fit a constant treatment effect
model, a linear interaction model and a RCS-3 model individualizing
absolute benefit predictions. RCS-4 and RCS-5 models were excluded. In
our simulations these methods were always outperformed by the simpler
approaches and were often overfitted. Finally, an adaptive approach with
only the 3 candidate models was also applied.

All considered methods provided similar fits, predicting increasing
benefits for patients with higher baseline risk predictions. All models
followed the evolution of the stratified estimates very closely. The
adaptive approach based on AIC selected the constant treatment effect
model. The constant treatment effect model had somewhat lower AIC
compared to the linear interaction model slightly worse cross-validated
discrimination (c-for-benefit 0.525 vs 0.526) and better cross-validated
calibration (ICI-for benefit 0.0104 vs 0.0115). In conclusion, a simpler
constant treatment effect model is adequate for predicting absolute
30-day mortality benefits of treatment with tPA in patients with with
acute MI.

@@

Risk-based approaches to predictive HTE estimate treatment benefit as a
function of baseline risk. A limitation of our study is that we assumed
treatment benefit to be a function of baseline risk in the majority of
the simulation scenarios. We also considered constant moderate and
strong treatment-related harms, applied on the absolute scale to expand
the range of scenarios in line with previous work {[}15{]}. In a limited
set of scenarios where we assumed the existence of true
treatment-covariate interactions, our conclusions remained unchanged.
Even though the average error rates increased for all the considered
methods, due to the miss-specification of the outcome model, the linear
interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased
results, espcially in the presence of moderate or strong
treatment-related harms. All the results of these simulations can be
found in Supplement, Section XX. Future simulation studies could explore
the effect of more extensive deviations from risk-based treatment
effects.

In our simulations we only focused on risk-based methods, using baseline
risk as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}16--19{]}. Many of these methods rely on
incorporating treatment-covariate interactions in the prediction of
benefit. An important caveat of such approaches is that they may be
prone to overfitting, thus exaggerating the magnitude of the predicted
benefits. In a wide range of simulation settings, a simpler risk
modeling approach was consistently better calibrated for benefit
compared to more complex treatment effect modelling approaches {[}5{]}.
Similarly, when SYNTAX score II, a model developed for identifying
patients with complex coronary artery disease that benefit more from
percutaneous coronary intervention or from coronary artery bypass
grafting was redeveloped using fewer treatment-covariate interactions
had better external performance compared to its predecessor{[}20,21{]}.

In conclusion, the linear interaction approach is a viable option with
smaller sample sizes and/or moderately performing risk prediction models
if we consider a non-constant relative treatment effect plausible. RCS-3
is a better option when non-monotonic deviations from a constant
relative treatment effect and/or substantial treatment-related harms are
anticipated. Increasing the complexity of the RCS models by increasing
the number of knots does not translate to improved benefit prediction.
Using AIC for model selection among the constant treatment effect, the
linear interaction and RCS-3 model is a viable option, especially with
larger sample size.

@@

\leavevmode\hypertarget{ref-Glasziou1995}{}%
{[}15{]} Glasziou PP, Irwig LM. An evidence based approach to
individualising treatment. BMJ 1995;311:1356--9.
\url{https://doi.org/10.1136/bmj.311.7016.1356}.

\leavevmode\hypertarget{ref-Athey2019}{}%
{[}16{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.

\leavevmode\hypertarget{ref-Lu2018}{}%
{[}17{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
treatment effect in observational data using random forest methods.
Journal of Computational and Graphical Statistics 2018;27:209--19.
\url{https://doi.org/10.1080/10618600.2017.1356325}.

\leavevmode\hypertarget{ref-Wager2018}{}%
{[}18{]} Wager S, Athey S. Estimation and inference of heterogeneous
treatment effects using random forests. Journal of the American
Statistical Association 2018;113:1228--42.
\url{https://doi.org/10.1080/01621459.2017.1319839}.

\leavevmode\hypertarget{ref-powers2018some}{}%
{[}19{]} Powers S, Qian J, Jung K, Schuler A, Shah NH, Hastie T, et al.
Some methods for heterogeneous treatment effect estimation in high
dimensions. Statistics in Medicine 2018;37:1767--87.

\leavevmode\hypertarget{ref-farooq2013anatomical}{}%
{[}20{]} Farooq V, Van Klaveren D, Steyerberg EW, Meliga E, Vergouwe Y,
Chieffo A, et al. Anatomical and clinical characteristics to guide
decision making between coronary artery bypass surgery and percutaneous
coronary intervention for individual patients: Development and
validation of syntax score ii. The Lancet 2013;381:639--50.

\leavevmode\hypertarget{ref-takahashi2020redevelopment}{}%
{[}21{]} Takahashi K, Serruys PW, Fuster V, Farkouh ME, Spertus JA,
Cohen DJ, et al. Redevelopment and validation of the syntax score ii to
individualise decision making between percutaneous and surgical
revascularisation in patients with complex coronary artery disease:
Secondary analysis of the multicentre randomised controlled syntaxes
trial with external cohort validation. The Lancet 2020;396:1399--412.
\end{cslreferences}

### 2022-02-02 submission/supplement.rmd
co-author comments

```{r, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)

@@

# Adaptive model selection frequencies

@@

\newpage

# Discrimination and calibration for benefit

@@

\newpage

# Strong relative treatment effect

@@

# Treatment interactions

@@

```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000. AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_positive.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000. AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_negative.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000. AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

# Empirical illustration

@@

treatmentArms <- gusto %>%
  group_by(treatment) %>%
  summarise(n = n())
```

\begin{equation}
P(outcome=1|X=x) = \text{expit}(lp(x)),
(\#eq:gusto1)
\end{equation}
where 
\begin{equation}
\begin{aligned}
lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + lsp(\text{pulse}, 50) +\\
&\beta_6 I(\text{pmi}=yes)+\beta_8 I(\text{miloc}=Anterior) + \\
&\beta_9 I(\text{miloc}=Other) + \gamma t_x
\end{aligned}
(\#eq:gusto2)
\end{equation}

>>>>

```{r, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(rms)

@@

\newpage
# Adaptive model selection frequencies

@@

\newpage
# Discrimination and calibration for benefit

@@

\newpage
# Strong relative treatment effect

@@

\newpage
# Treatment interactions

@@

```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring treatment were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_positive.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring the control were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_negative.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions 2 favoring treatment and 2 favoring the control were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

\newpage
# Empirical illustration

@@

treatmentArms <- gusto %>%
  group_by(treatment) %>%
  summarise(n = n())

prediction <- glm(
   outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
   data = gusto,
   family = "binomial",
   maxit = 99
)

predictionCoefficients <- coef(prediction)
predictionCoefficients <- predictionCoefficients[c(1, 3:12, 2)]
```

\begin{equation}
P(outcome=1|X=x) = \text{expit}(lp(x)),
(\#eq:gusto1)
\end{equation}
where 
\begin{equation}
\begin{aligned}
lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + \beta_6 \text{pulse}+\\
&\beta_7 max(\text{pulse - 50, 0}) + \beta_8 I(\text{pmi}=yes)+\\
&\beta_9 I(\text{miloc}=Anterior) + \beta_9 I(\text{miloc}=Other) +\\
&\gamma\times\text{treatment}
\end{aligned}
(\#eq:gusto2)
\end{equation}

and $expit(x) = \frac{e^x}{1+e^x}$

```{r, echo=FALSE}
predictionSummary <- summary(prediction)
# predictionSummaryCoef <- tibble(predictionSummary$coefficients) %>%
#   rownames_to_column("Variable")

predictionCoefficientsTable <- predictionSummary$coefficients %>%
  as.data.frame() %>%
  mutate(
    rownames = c(
      "Intercept",
      "Age",
      "Killip class = II",
      "Killip class = III",
      "Killip class = IV",
      "Systolic blood pressure",
      "Pulse rate (1)",
      "Pulse rate (2)",
      "Previous MI (yes)",
      "MI location (Other)",
      "MI location (Anterior)",
      "Treatment"
    )
  ) %>%
  rownames_to_column("Variable") %>%
  select(-Variable) %>%
  select(rownames, everything())

colnames(predictionCoefficientsTable) <- c(
  "Variable",
  "Estimate",
  "stderror",
  "zvalue",
  "pvalue"
)

kableExtra::kbl(
  predictionCoefficientsTable,
  format = "latex",
  escape = FALSE,
  longtable = TRUE,
  align = c("l", rep("r", 4)),
  booktabs = TRUE,
  caption = "Coefficients of the prediction model for 30-day mortality, based on the data from GUSTO-I trial.",
  digits = 3,
  linesep = "",
  format.args = list(
    big.mar      = ",",
    decimal.mark = "."
  )
) %>%
  kableExtra::kable_styling(font_size = 7)
```
